---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SoilGeoStats


The goal of SoilGeoStats is to provide a reproducible, multi-tiered spatial interpolation pipeline tailored for official topsoil point-observation datasets.

This is a Data Analysis package structured around a regional environmental case study in Bavaria, Germany, and inspired by a Spatial Statistics course project carried out in a previous semester [Geostatistical Analysis of SOC](https://github.com/adebolaadedayo/Geostatistical-Analysis-of-SOC) using ArcGIS.

It combines automated spatial data cleaning with both deterministic modeling and advanced multivariate geostatistical analysis. While initialized for a Bavarian environmental assessment, this package is completely generic and can be dynamically applied to topsoil data across other geographic locations.

## Target Users

The target users of this package are spatial data analysts, environmental scientists, soil ecologists, and researchers who work with soil point datasets and need a quick, reproducible pipeline to clip, validate, and carry out geostatistical interpolation.

## Dataset and Source

* Topsoil Observations: The enclosed point dataset contains topsoil observations (0–20 cm depth) extracted from the official [LUCAS 2018 Topsoil data](https://esdac.jrc.ec.europa.eu/content/lucas-2018-topsoil-data) compiled by the European Soil Data Centre (ESDAC). The dataset covers the European Union and the United Kingdom, containing attributes for elevation, pH, organic carbon content (OC), phosphorus, potassium, and nitrogen (N).

* Regional Boundary: The package accommodates an optional vector boundary polygon shapefile of Bavaria obtained from the Database of [Global Administrative Areas (GADM)](https://gadm.org/data.html) to dynamically isolate regional study areas.

* Coordinate Reference Systems: The raw LUCAS point dataset is provided in unprojected WGS84 (EPSG: 4326) coordinates, while the GADM boundary shapefile of Bavaria uses the projected ETRS89 / UTM zone 32N (EPSG: 25832) reference system. The package automatically reconciles this CRS mismatch during the ingestion pipeline.

## Analysis Workflow

The analytical pipeline enables a user to evaluate soil properties sequentially through six distinct tiers:

* Data Validation and Geometric Clipping: Ingests tabular CSV profiles and an optional regional polygon mask, executing an automated spatial point-in-polygon clip.

* Distance Metrics: Builds geographic distance matrices to evaluate sampling intervals alongside exploratory data analysis summaries.

* Deterministic Interpolation: Computes a local spatial prediction matrix using Inverse Distance Weighting (IDW).

* Regression Kriging: Maximizes prediction accuracy by pairing Ordinary Least Squares (OLS) environmental trends (using Nitrogen and Elevation covariates) with spatial residual interpolation.

* Data Visualization: Transforms S3 prediction grids into boundary-clipped continuous map tile visualizations via ggplot2.

* Performance Validation: Evaluates model accuracy using iterative Leave-One-Out Cross-Validation (LOOCV) to output Mean Error (ME) for prediction bias and Root Mean Square Error (RMSE) for modeling precision.

## Installation

You can install the development version of SoilGeoStats from GitHub using the `remotes` package:

``` {r, eval = FALSE}
# Install the development version from GitHub
# install.packages("remotes")
remotes::install_github("adebolaadedayo/SoilGeoStats")
```

## Demonstration Example

Here is a complete, reproducible walkthrough demonstrating how to initialize the data, execute the dual-interpolation engines, evaluate cross-validation accuracy metrics, and plot the resulting continuous environmental surfaces:

```{r example}
library(SoilGeoStats)
library(sf)
library(ggplot2)

# Step 1 & 2: Ingest data using built-in package system defaults and verify spacing
soil_points <- load_and_validate_data()
spatial_EDA <- calculate_spatial_distances(spatial_data = soil_points, target_col = "OC")

# View exploratory metrics calculated by the pipeline
print(spatial_EDA$total_points)
print(spatial_EDA$minimum_sampling_interval_meters)

# Step 3 & 4: Execute Interpolation Engines (Instantiates our custom S3 Classes)
idw_surface <- interpolate_idw_surface(spatial_data = soil_points, target_col = "OC")
rk_surface  <- interpolate_regression_kriging(spatial_data = soil_points, target_col = "OC", covariate_cols = c("N", "Elev"))

# Step 5: S3 Method Validation - Print Object Summary Metadata
print(idw_surface)

# Step 6: S3 Method Validation - Execute Automated Leave-One-Out Cross-Validation Math
summary(idw_surface)
summary(rk_surface)
```

## Map Visualizations

By using the custom S3 plot() method, you can pass the background boundary file directly to generate beautifully clipped continuous map tile surfaces:

```{r}
# Load the background polygon file path from the package system
bavaria_layer <- sf::st_read(system.file("extdata", "shape_file/Bav_Boundary.shp", package = "SoilGeoStats"), quiet = TRUE)

# Render the continuous maps using our custom S3 plot methods
plot(idw_surface, boundary_data = bavaria_layer)
plot(rk_surface, boundary_data = bavaria_layer)
```


