---
title: "Geostatistical Modelling of Soil Organic Carbon in Bavaria"
author: "<center><medium>Adebola David Adedayo</medium></center>"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geostatistical Modelling of Soil Organic Carbon in Bavaria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```


# Introduction and Background

This vignette presents a reproducible geostatistical data pipeline evaluating Soil Organic Carbon (OC) distribution across Bavaria, Germany. Using a dataset obtained from the official [LUCAS 2018 Topsoil data](https://esdac.jrc.ec.europa.eu/content/lucas-2018-topsoil-data) compiled by the European Soil Data Centre (ESDAC), this package will be useful to spatial data analysts, environmental scientists, soil ecologists, and researchers who work with soil point datasets and need a quick, reproducible pipeline to clip, validate, and carry out geostatistical interpolation.

The study workflow showcases standard GIS methods upgraded into an object-oriented R programming framework utilizing S3 method dispatching.

```{r}
library(SoilGeoStats)
library(sf)
library(ggplot2)
```

## I. Ingestion, Validation, and Spatial Clipping

The first stage of the analysis pipeline ingests unprojected WGS84 point records and matches them against a regional vector boundary shapefile of Bavaria (EPSG: 25832) obtained from [GADM](https://gadm.org/data.html). The function automatically resolves coordinate mismatches, carries out a point-in-polygon intersection clip, and appends the resulting coordinate values.

```{r}
# Ingest and spatially process the dataset using system file paths
soil_data_raw <- load_and_validate_data()

# Filter to essential columns only
soil_data_clipped <- dplyr::select(soil_data_raw, OC, N, Elev, geometry)

# View spatial features layout
print(head(soil_data_clipped))
```

**Column Definitions:**

`OC` (Organic Carbon Content): the primary target variable,representing the mass fraction of organic carbon in the topsoil (g/kg).

`N` (Nitrogen): an environmental covariate representing the total nitrogen content (g/kg), highly correlated with organic matter accumulation.

`Elev` (Elevation): A topographic covariate representing altitude above sea level (meters).

`geometry` (Spatial Features): The metric coordinate point layouts in the ETRS89 / UTM zone 32N grid system.

### Visualizing the Spatial Sample Distribution

Before executing the interpolation surfaces, the cleaned point observation locations are mapped to evaluate how evenly the physical soil samples are dispersed across the Bavarian state boundary:

```{r point_map, fig.width=7, fig.height=5}

# Load Bavaria boundary polygon
bavaria_mask <- sf::st_read(system.file("extdata", "shape_file/Bav_Boundary.shp", package = "SoilGeoStats"), quiet = TRUE)

# Plot point distributions
ggplot() +
  geom_sf(data = bavaria_mask, fill = "gray95", color = "gray60") +
  geom_sf(data = soil_data_clipped, aes(color = OC), size = 1.5, alpha = 0.8) +
  scale_color_viridis_c(name = "Soil Organic Carbon\n(g/kg)", option = "viridis") +
  coord_sf(datum = sf::st_crs(25832)) +
  labs(title = "LUCAS 2018 Topsoil Sample Point Distribution",
       subtitle = "Spatially Clipped to Bavaria Study Area") +
  theme_minimal() +
  theme(legend.position = "right")

```


## II. Exploratory Spatial Data Analysis

Before running interpolations, a spatial distance analysis is executed to compute minimum and maximum point-pair bounds alongside basic variable summary metrics (such as mean, median, variance).

```{r}
# Execute spatial distance metric profiling
eda_metrics <- calculate_spatial_distances(spatial_data = soil_data_clipped, target_col = "OC")

# Display compiled summary metrics
cat("Total Observations Inside Mask  :", eda_metrics$total_points, "points\n")
cat("Minimum Point Spacing Interval  :", round(eda_metrics$minimum_sampling_interval_meters / 1000, 2), "km\n")
cat("Maximum Study Extent Diameter   :", round(eda_metrics$maximum_sampling_interval_meters / 1000, 2), "km\n")
cat("Sample Mean of Target Column:", eda_metrics$variable_mean, "\n")
cat("Sample Median of Target Column:", eda_metrics$variable_median, "\n")
cat("Sample Variance of Target Column:", eda_metrics$variable_variance, "\n")
```

**Statistical Significance of Results:**

- Total Observations (`total_points`): Represents the observation sample size available within the geographic region. A robust point count provides the data density required to build statistically sound interpolation weights.

- Minimum Point Spacing Interval (`minimum_sampling_interval_meters`): Signifies the absolute finest spatial resolution of the raw data collection grid. This metric tells us how close the nearest neighbor points are.

- Maximum Study Extent Diameter (`maximum_sampling_interval_meters`): Measures the absolute maximum geographic distance stretching across the study polygon. It is critical for constraining interpolation search radii.

- Sample Mean (`variable_mean`): This provides the global mathematical baseline of Topsoil Organic Carbon across Bavaria.

- Sample Median (`variable_median`): This identifies the exact midpoint of the data distribution. Comparing the mean directly to the median allows us to detect statistical skewness. If the mean is significantly higher than the median, it indicates a right-skewed distribution, meaning the landscape has localized "hotspots" of high carbon storage among lower baseline agricultural values.

- Sample Variance (`variable_variance`): Measures the global dispersion and heterogeneity of the Soil Organic Carbon data. Higher variance values signal an environmentally complex region with significant localized shifts, highlighting why advanced geostatistical modeling (like Regression Kriging) is necessary over simple averages.


## III. Spatial Interpolation (S3 Object Construction)

The deterministic and hybrid geostatistical modeling engines are executed. The functions automatically output instances of the custom S3 `spatial_grid` class.

```{r}
# 1. Deterministic Inverse Distance Weighting Interpolation
idw_grid_obj <- interpolate_idw_surface(spatial_data = soil_data_clipped, target_col = "OC", power = 2.0)

# 2. Multi-Variable Regression Kriging (Using Nitrogen and Elevation Covariates)
rk_grid_obj  <- interpolate_regression_kriging(spatial_data = soil_data_clipped, 
                                                target_col = "OC", 
                                                covariate_cols = c("N", "Elev"))
```

## IV. S3 Methods (Print and Cross-Validation)

By simply calling standard R commands like `print()` and `summary()`, R automatically resolves the custom S3 hierarchy stamp, routing the data directly to the created leave-one-out cross-validation (LOOCV) calculations.

Object Metadata:
```{r}
print(idw_grid_obj) #View output of IDW interpolation
print(rk_grid_obj) #View output of regression kriging
```

Interpolation Model Accuracy Assessments (LOOCV Results):
```{r}
summary(idw_grid_obj) #Accuracy metrics of IDW interpolation
summary(rk_grid_obj) #Accuracy metrics of regression kriging
```

## V. Cartographic Visualizations

Finally, the custom S3 `plot()` method is explored. By integrating the study region's polygon file path directly into the generic function, it automatically generates continuous map tile surfaces clipped cleanly to study region.

```{r}
# Extract internal shape path for vector outlines
bavaria_mask <- sf::st_read(system.file("extdata", "shape_file/Bav_Boundary.shp", package = "SoilGeoStats"), quiet = TRUE)

# Generate custom continuous map objects
map_idw <- plot(idw_grid_obj, boundary_data = bavaria_mask) #IDW surface
map_rk  <- plot(rk_grid_obj, boundary_data = bavaria_mask) #Regression Krigging Surface

# Render the final cartographic surfaces
print(map_idw)
print(map_rk)
```

